Bayesian Learning of Degenerate Linear~Gaussian~State~Space~Models using~Markov~Chain~Monte~Carlo

### 2015-09-22 notes/paper-degenerate-linear-systems.tex
Edited.

\usepackage{amsmath}
\usepackage{amsfonts}
%\usepackage{IEEEtrantools}
\usepackage{algorithm}
\usepackage{algorithmic}
% \usepackage{graphicx}
\usepackage{hyperref}
\usepackage{subfig}
\usepackage{cite}

@@

The paper is structured as follows. In section~\ref{sec:linear_gaussian_models} we review the Gibbs sampler for basic linear Gaussian state space systems, focusing on the transition model. In section~\ref{sec:degenerate_transition_models} we describe modifications for degenerate transitions. Simulations on a toy model, and a motion capture application are presented in section~\ref{sec:applications}.

@@

The conjugate prior for $\lgtm, \lgtv$ is a matrix normal--inverse Wishart distribution \cite{Wills2012},
%
\begin{align}
 \lgtv &\sim \iwishartdist{\priordof}{\priorscalematrix} \\
 \lgtm | \lgtv &\sim \matrixnormaldist{\priormeanmatrix}{\lgtv}{\priorcolumnvariance}     ,
\end{align}
%
with the following density function,
%
\begin{equation}
 \den(\lgtm, \lgtv) =  \den(\lgtm | \lgtv) \den(\lgtv)
\end{equation}
%
\begin{align}
\den(\lgtv) &= \frac{ \determ{\priorscalematrix}^{\frac{\priordof}{2}} }{ 2^{\frac{\priordof}{2}} \gammafun\left(\frac{\priordof}{2}\right) } \determ{\lgtv}^{-\frac{\priordof+\lsd+1}{2}} \exp\left( -\half \trace\left[\lgtv\inv\priorscalematrix\right] \right) \\
\den(\lgtm | \lgtv) &= \determ{2 \pi \lgtv}^{-\half} \determ{2 \pi \priorcolumnvariance}^{-\half} \exp\left(-\half \trace\left[ \lgtv\inv (\lgtm-\priormeanmatrix) \priorcolumnvariance\inv (\lgtm-\priormeanmatrix)\tr \right] \right) 
\end{align}
%
$\priordof\in\reals, \priordof>\lsd-1$. $\priorscalematrix$ and $\priorcolumnvariance$ are $\lsd\times\lsd$ positive definite matrices. $\priormeanmatrix \in \reals^{\lsd\times\lsd}$.

@@

An appropriate Gibbs sampler can be constructed by sampling alternately from the following conditional posterior distributions,
%
\begin{IEEEeqnarray}{l}
 \postden(\ls{1:\timax}|\lgtm, \lgtv) \nonumber \\
 \postden(\lgtm, \lgtv| \ls{1:\timax}) \nonumber      .
\end{IEEEeqnarray}
%
The distribution of the sampled values will then converge to the target posterior distribution \cite{Roberts1994}. The state conditional $\postden(\ls{1:\timax}|\lgtm, \lgtv)$ can be sampled using the forward-filtering-backward-sampling algorithm \cite{Chib1996,Wills2012}, which works by first running a Kalman filter forwards through the data, then running backwards in time sampling each state conditional on those in the future. The parameter conditional may also be sampled directly. This depends on the state sequence probability,
%
\begin{IEEEeqnarray}{rCl}
 \IEEEeqnarraymulticol{3}{l}{ \den(\ls{1:\timax}|\lgtm,\lgtv) \propto \prod_{\ti=2}^{\timax} \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) } \\
 \quad\quad & \propto & \determ{\lgtv}^{-\frac{\timax-1}{2}} \exp\left( -\half \sum_{\ti=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})\tr \lgtv\inv (\ls{\ti}-\lgtm\ls{\ti-1}) \right)      .
\end{IEEEeqnarray}

It is straightforward to show that the required conditional distribution is also a matrix normal--inverse Wishart distribution \cite{Wills2012},
%
\begin{IEEEeqnarray}{rCl}
 \lgtv | \ls{1:\timax} &\sim& \iwishartdist{\postdof}{\postscalematrix} \\
 \lgtm | \lgtv, \ls{1:\timax} &\sim& \matrixnormaldist{\postmeanmatrix}{\lgtv}{\postcolumnvariance}     ,
\end{IEEEeqnarray}
%
with the following updated hyperparameters,
%
\begin{align}
 \postcolumnvariance\inv                 &= \priorcolumnvariance\inv + \suffstats{1} \\
 \postmeanmatrix \postcolumnvariance\inv &= \priormeanmatrix \priorcolumnvariance\inv + \suffstats{2}\\
 \postdof                                &= \priordof + \suffstats{0} \\
 \postscalematrix                        &= \priorscalematrix + \suffstats{3} + \priormeanmatrix \priorcolumnvariance\inv \priormeanmatrix\tr - \postmeanmatrix \postcolumnvariance\inv \postmeanmatrix\tr    ,
\end{align}
%
in which the following sufficient statistics are used,
%
\begin{align}
 \suffstats{0} &= \timax - 1 \\
 \suffstats{1} &= \sum_{t=2}^{\timax} \ls{\ti-1}\ls{\ti-1}\tr \\
 \suffstats{2} &= \sum_{t=2}^{\timax} \ls{\ti}\ls{\ti-1}\tr \\
 \suffstats{3} &= \sum_{t=2}^{\timax} \ls{\ti}\ls{\ti}\tr      .
\end{align}

@@

A further factorisation will also be useful, which we will refer to as the Givens decomposition (on account of its method of computation). For the rank-$\rk$ positive semi-definite covariance matrix $\lgtp$,
%
\begin{IEEEeqnarray}{rClCl}
 \lgtv &=& \tvrot \tvfull \tvrot\tr &=& \begin{bmatrix}\tvrot & \tvrotorth\end{bmatrix} \begin{bmatrix}\tvfull & \zmat \\ \zmat & \zmat\end{bmatrix} \begin{bmatrix}\tvrot\tr \\ \tvrotorth\tr\end{bmatrix} \label{eq:givens_decomposition}
\end{IEEEeqnarray}
%
where,
%
$\tvfull$ is a $\rk\times\rk$ positive definite matrix and $\tvrot$ is a $\lsd\times\rk$ matrix of orthonormal vectors subject to certain constraints. The transformation from $\lgtv$ to $\{\tvfull, \tvrot\}$ is also bijective. See appendix~\ref{app:givens-factorisation} for details.

@@

An appropriate prior distribution for the degenerate transition covariance matrix may be defined by setting the degrees of freedom of an inverse Wishart distribution to the desired rank, $\priordof=\rk$. This results in a singular inverse Wishart distribution over the rank-$\rk$ positive semi-definite matrices \cite{Diaz-Garcia2006},
%
\begin{IEEEeqnarray}{rCl}
 \lgtv|\rk &\sim& \iwishartdist{\rk}{\priorscalematrixrank{\rk}}     .
\end{IEEEeqnarray}
%
The density (with respect to the Lesbegue measure on the distinct elements) is,
%
\begin{IEEEeqnarray}{rCl}
 \den(\lgtv|\rk) &=& \frac{ \determ{\priorscalematrixrank{\rk}}^{\frac{\rk}{2}} }{ 2^{\half\rk\lsd} \pi^{\half\rk(\lsd-\rk)} \gammafun[r]\left(\frac{\rk}{2}\right) } \determ{\tvval}^{-\half(3\lsd-\rk+1)} \exp\left( -\half \lgtv\pinv \priorscalematrixrank{\rk} \right)    ,
\end{IEEEeqnarray}
%
in which $\lgtv\pinv$ denotes the Moore-Penrose pseudoinverse of $\lgtv$. A prior estimate of $\lgtv$ is $\frac{1}{\rk}\priorscalematrixrank{\rk}$. Thus, in order to maintain consistency between the different ranks, we should set $\priorscalematrixrank{\rk} = \rk \priorscalematrixbase$, where $\priorscalematrixbase$ is an estimate of $\lgtv$.

@@

\subsubsection{Likelihood and Subspace Constraints}
There is no transition density over $\reals^{\lsd}$ for the degenerate model. However, a density does exist with respect to an appropriate measure on the reachable subspace, which provides us with a likelihood function \cite{Diaz-Garcia2006}. We can derive this from the underlying state transition equation,
%
\begin{IEEEeqnarray}{rCl}
 \ls{\ti} &=& \lgtm \ls{\ti-1} + \tvrot \tvfull\msqrt \tn{\ti} \nonumber
 \end{IEEEeqnarray}
 \begin{IEEEeqnarray}{rCl}
 \Rightarrow \tvrot\tr(\ls{\ti}-\lgtm\ls{\ti-1}) &\sim& \normaldist{\zmat}{\tvfull} \nonumber
\end{IEEEeqnarray}
\begin{IEEEeqnarray}{rCl}
 \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) &\propto& \determ{\tvfull}^{-\half} \exp\left(-\half (\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \tvfull\inv \tvrot\tr (\ls{\ti}-\lgtm\ls{\ti-1}) \right)     .
\end{IEEEeqnarray}
%
The subspace over which this is defined is,
%
\begin{IEEEeqnarray}{rCl}
 \{\ls{\ti} : \ls{\ti} = \lgtm \ls{\ti-1} + \tvrot z, z \in \reals^{\rk}\}     .
\end{IEEEeqnarray}

The likelihood function for the entire state trajectory is thus,
%
\begin{align}
 \den(\ls{1:\timax}|\lgtm, \lgtv) &= \den(\ls{0}) \prod_{t=2}^{\timax} \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) \nonumber \\
 &\propto \determ{\tvfull}^{-\half(\timax-1)} \exp\left( -\half \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \tvfull\inv \tvrot\tr (\ls{\ti}-\lgtm\ls{\ti-1}) \right) \nonumber \\
 &= \determ{\tvfull}^{-\half(\timax-1)} \exp\left(-\half\trace\left[ \tvfull\inv \tvrot\tr \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})(\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \right]\right) \nonumber \\
 &= \determ{\tvfull}^{-\half\suffstats{0}} \exp\left(-\half\trace\left[ \tvfull\inv \left( \tvrot\tr \lgtm \suffstats{1} \lgtm\tr \tvrot - \tvrot\tr \lgtm \suffstats{2}\tr \tvrot - \tvrot\tr \suffstats{2} \lgtm\tr \tvrot + \tvrot\tr \suffstats{3} \tvrot \right) \right]\right) \\
 &= \determ{\tvfull}^{-\half\suffstats{0}} \exp\left(-\half\trace\left[ \tvfull\inv \left( \lgtmrot \suffstats{1} \lgtmrot\tr - \lgtmrot \suffstats{2}\tr \tvrot - \tvrot\tr \suffstats{2} \lgtmrot\tr + \tvrot\tr \suffstats{3} \tvrot \right) \right]\right)
\end{align}
%
in which the sufficient statistics are as before and,
%
\begin{equation}
 \lgtmrot = \tvrot\tr \lgtm      .
\end{equation}

@@

\subsubsection{Sampling the Conditional}
Using the constrained likelihood function and the transformed prior, we now obtain the required posterior conditional distribution $\postden(\lgtmrot,\tvfull|\lgtmorth,\tvrot,\ls{1:\timax})$. This follows the same algebraic steps as in the full rank case and leads to a similar matrix normal--inverse Wishart distribution,
%
\begin{align}
 \tvfull | \tvrot, \ls{1:\timax} &\sim \iwishartdist{\postdof}{\postscalematrix} \\
 \lgtmrot | \tvfull, \tvrot, \ls{1:\timax} &\sim \matrixnormaldist{\postmeanmatrix}{\lgtv}{\postcolumnvariance}     ,
\end{align}
%
with the following updated hyperparameters,
%
\begin{align}
 \postcolumnvariance\inv                 &= \priorcolumnvariance\inv + \suffstats{1} \\
 \postmeanmatrix \postcolumnvariance\inv &= \tvrot\tr\left( \priormeanmatrix \priorcolumnvariance\inv + \suffstats{2} \right) \\
 \postdof                                &= \rk + \suffstats{0} \\
 \postscalematrix                        &= (\tvrot\tr\priorscalematrix\inv\tvrot)\inv + \tvrot\tr\left( \suffstats{3} + \priormeanmatrix \priorcolumnvariance\inv \priormeanmatrix\tr \right)\tvrot - \postmeanmatrix \postcolumnvariance\inv \postmeanmatrix\tr
\end{align}

@@

\section{Reversible Jump for Rank Learning}

@@

The acceptance probability for this move and its reverse respectively are \cite{Green1995},
%
\begin{IEEEeqnarray}{rCl}
 \mhap(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew) &=& \min\left\{1, \beta \right\} \\
 \mhap(\lgtv\mcnew\to\lgtv,\eval{\rk+1},\evec{\rk+1}) &=& \min\left\{1, \beta\inv \right\} 
\end{IEEEeqnarray}
%
\begin{IEEEeqnarray}{rCl}
 \beta &=& \frac{ \postden(\lgtm, \lgtv\mcnew, \rk+1) \jac(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew) }{ \postden(\lgtm,\lgtv,\rk) \ppslden{\eval{}}(\eval{\rk+1}|\lgtv) \ppslden{\evec{}}(\evec{\rk+1}|\lgtv) } \times \frac{P(\rk+1\to\rk)}{P(\rk\to\rk+1)}     ,
\end{IEEEeqnarray}
%
where $\jac(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew)$ is the Jacobian of the transformation and $P(\rk+1\to\rk)$ is the probability of performing a move which changes the rank from $\rk$ to $\rk+1$.

The Jacobian of the eigendecomposition for a positive semi-definite matrix is \cite{Diaz-Garcia2006},
%
\begin{IEEEeqnarray}{rCl}
 \jac(\tvval,\tvvec\to\lgtv) &=& 2^{-\rk} \times \prod_{i=1}^{\rk}\eval{i}^{\lsd-\rk} \times \prod_{j=1}^{r}\prod_{i=1}^{j-1}(\eval{i}-\eval{j})      .
\end{IEEEeqnarray}
%
Repeating this for $\lgtv\mcnew$ and taking the ratio of the two, we obtain,
%
\begin{IEEEeqnarray}{rCl}
 \jac(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew) &=& 2^{-1} \eval{\rk+1}^{\lsd-\rk-1} \prod_{i=1}^{\rk}\eval{i}^{-1} \prod_{i=1}^{\rk}(\eval{i}-\eval{\rk+1})     .
\end{IEEEeqnarray}

\section{Simulations}

@@

The hyperparameters were set to generally uninformative values. For the full rank model,
%
\begin{IEEEeqnarray}{rCl}
 \priordof &=& \lsd - 1 \\
 \priorscalematrix &=& \priordof \idmat \\
 \priormeanmatrix &=& \zmat \\
 \priorcolumnvariance &=& 100 \idmat      .
\end{IEEEeqnarray}
%
Similarly, for the degenerate model,
%
\begin{IEEEeqnarray}{rCl}
 \priorscalematrixbase &=& \idmat \\
 \priormeanmatrix &=& \zmat \\
 \priorcolumnvariance &=& 100 \idmat \\
 \priortypval &=& 1     .
\end{IEEEeqnarray}

@@

\begin{figure}
 \centering
 \subfloat[]{\includegraphics[width=0.9\columnwidth]{figures/toy-basic-R-acf.pdf} \label{subf:toy-R-acf:basic}} \\
 \subfloat[]{\includegraphics[width=0.9\columnwidth]{figures/toy-degenerate-R-acf.pdf} \label{subf:toy-R-acf:degenerate}}
 \caption{MCMC autocorrelation plots for $\xi_{\ob{}}$ for the full rank transition model \protect\subref{subf:toy-R-acf:basic} and the degenerate transition model \protect\subref{subf:toy-R-acf:degenerate}.}
 \label{fig:toy-R-acf}
\end{figure}

@@

The hyperparameters are set to the same uninformative values as for the toy model, with the following exception,
%
\begin{IEEEeqnarray}{rCl}
 \priorscalematrix &=& 0.001 \priordof \idmat \\
 \priorscalematrixbase &=& 0.001 \idmat \\
 \priortypval &=& 0.01     .
\end{IEEEeqnarray}

@@

\bibliographystyle{IEEEtran}
% \bibliography{/home/pete/Dropbox/PhD/bibliographies/Cleanbib}
\bibliography{/home/pete/Dropbox/PhD/bibliographies/OTbib}
\end{document}

>>>>

\usepackage{amsmath}
\usepackage{amsfonts}
%\usepackage{IEEEtrantools}
\usepackage{algorithm}
\usepackage{algorithmic}
% \usepackage{graphicx}
\usepackage{hyperref}
\usepackage[caption=false]{subfig}
\usepackage{cite}

@@

The paper is structured as follows. In section~\ref{sec:linear_gaussian_models} we review the Gibbs sampler for basic linear Gaussian state space systems, focusing on the transition model. In section~\ref{sec:degenerate_transition_models} we describe modifications for degenerate transitions, and in section~\ref{sec:rank_learning} we show how the rank of the transition covariance may be learned. Simulations on a toy model, and a motion capture application are presented in section~\ref{sec:simulations}.

@@

The conjugate prior for $\lgtm, \lgtv$ is a matrix normal--inverse Wishart distribution \cite{Wills2012},
%
\begin{align}
 \lgtv &\sim \iwishartdist{\priordof}{\priorscalematrix} \\
 \lgtm | \lgtv &\sim \matrixnormaldist{\priormeanmatrix}{\lgtv}{\priorcolumnvariance}     ,
\end{align}
%
with the following density function,
%
\begin{equation}
 \den(\lgtm, \lgtv) =  \den(\lgtm | \lgtv) \den(\lgtv)
\end{equation}
%
\begin{IEEEeqnarray}{rCl}
\den(\lgtv) &=& \frac{ \determ{\priorscalematrix}^{\frac{\priordof}{2}} }{ 2^{\frac{\priordof}{2}} \gammafun\left(\frac{\priordof}{2}\right) } \determ{\lgtv}^{-\frac{\priordof+\lsd+1}{2}} \exp\left( -\half \trace\left[\lgtv\inv\priorscalematrix\right] \right) \nonumber \\
\\
\den(\lgtm | \lgtv) &=& \determ{2 \pi \lgtv}^{-\half} \determ{2 \pi \priorcolumnvariance}^{-\half} \nonumber \\
& & \times \exp\left(-\half \trace\left[ \lgtv\inv (\lgtm-\priormeanmatrix) \priorcolumnvariance\inv (\lgtm-\priormeanmatrix)\tr \right] \right) \nonumber \\
\end{IEEEeqnarray}
%
$\priordof\in\reals, \priordof>\lsd-1$. $\priorscalematrix$ and $\priorcolumnvariance$ are $\lsd\times\lsd$ positive definite matrices. $\priormeanmatrix \in \reals^{\lsd\times\lsd}$.

@@

An appropriate Gibbs sampler can be constructed by sampling alternately from the following conditional posterior distributions,
%
\begin{IEEEeqnarray}{l}
 \postden(\ls{1:\timax}|\lgtm, \lgtv) \nonumber \\
 \postden(\lgtm, \lgtv| \ls{1:\timax}) \nonumber      .
\end{IEEEeqnarray}
%
The distribution of the sampled values will then converge to the target posterior distribution \cite{Roberts1994}. The state conditional $\postden(\ls{1:\timax}|\lgtm, \lgtv)$ can be sampled using the forward-filtering-backward-sampling algorithm \cite{Chib1996,Wills2012}, which works by first running a Kalman filter forwards through the data, then running backwards in time sampling each state conditional on those in the future. The parameter conditional may also be sampled directly. This depends on the state sequence probability,
%
\begin{IEEEeqnarray}{rCl}
 \IEEEeqnarraymulticol{3}{l}{ \den(\ls{1:\timax}|\lgtm,\lgtv) \propto \prod_{\ti=2}^{\timax} \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) } \\
 & \propto & \determ{\lgtv}^{-\frac{\timax-1}{2}} \exp\left( -\half \sum_{\ti=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})\tr \lgtv\inv (\ls{\ti}-\lgtm\ls{\ti-1}) \right) \nonumber      .
\end{IEEEeqnarray}

It is straightforward to show that the required conditional distribution is also a matrix normal--inverse Wishart distribution \cite{Wills2012},
%
\begin{IEEEeqnarray}{rCl}
 \lgtv | \ls{1:\timax} &\sim& \iwishartdist{\postdof}{\postscalematrix} \\
 \lgtm | \lgtv, \ls{1:\timax} &\sim& \matrixnormaldist{\postmeanmatrix}{\lgtv}{\postcolumnvariance}     ,
\end{IEEEeqnarray}
%
with the following updated hyperparameters,
%
\begin{align}
 \postcolumnvariance\inv                 &= \priorcolumnvariance\inv + \suffstats{1} \nonumber \\
 \postmeanmatrix \postcolumnvariance\inv &= \priormeanmatrix \priorcolumnvariance\inv + \suffstats{2} \nonumber \\
 \postdof                                &= \priordof + \suffstats{0} \nonumber \\
 \postscalematrix                        &= \priorscalematrix + \suffstats{3} + \priormeanmatrix \priorcolumnvariance\inv \priormeanmatrix\tr - \postmeanmatrix \postcolumnvariance\inv \postmeanmatrix\tr    ,
\end{align}
%
in which the following sufficient statistics are used,
%
\begin{IEEEeqnarray}{rClCrCl}
 \suffstats{0} &=& \timax - 1 & \qquad & \suffstats{1} &=& \sum_{t=2}^{\timax} \ls{\ti-1}\ls{\ti-1}\tr \nonumber \\
 \suffstats{2} &=& \sum_{t=2}^{\timax} \ls{\ti}\ls{\ti-1}\tr & \qquad & \suffstats{3} &=& \sum_{t=2}^{\timax} \ls{\ti}\ls{\ti}\tr      .
\end{IEEEeqnarray}

@@

A further factorisation will also be useful, which we will refer to as the Givens decomposition (on account of its method of computation). For the rank-$\rk$ positive semi-definite covariance matrix $\lgtp$,
%
\begin{IEEEeqnarray}{rCl}
 \lgtv &=& \tvrot \tvfull \tvrot\tr %&=& \begin{bmatrix}\tvrot & \tvrotorth\end{bmatrix} \begin{bmatrix}\tvfull & \zmat \\ \zmat & \zmat\end{bmatrix} \begin{bmatrix}\tvrot\tr \\ \tvrotorth\tr\end{bmatrix} \label{eq:givens_decomposition}
\end{IEEEeqnarray}
%
where,
%
$\tvfull$ is a $\rk\times\rk$ positive definite matrix and $\tvrot$ is a $\lsd\times\rk$ matrix of orthonormal vectors subject to certain constraints. The transformation from $\lgtv$ to $\{\tvfull, \tvrot\}$ is also bijective. See appendix~\ref{app:givens-factorisation} for details.

@@

An appropriate prior distribution for the degenerate transition covariance matrix may be defined by setting the degrees of freedom of an inverse Wishart distribution to the desired rank, $\priordof=\rk$. This results in a singular inverse Wishart distribution over the rank-$\rk$ positive semi-definite matrices \cite{Diaz-Garcia2006},
%
\begin{IEEEeqnarray}{rCl}
 \lgtv|\rk &\sim& \iwishartdist{\rk}{\priorscalematrixrank{\rk}}     .
\end{IEEEeqnarray}
%
The density (with respect to the Lesbegue measure on the distinct elements) is,
%
\begin{IEEEeqnarray}{rCl}
 \den(\lgtv|\rk) &=& \frac{ \determ{\priorscalematrixrank{\rk}}^{\frac{\rk}{2}} }{ 2^{\half\rk\lsd} \pi^{\half\rk(\lsd-\rk)} \gammafun[r]\left(\frac{\rk}{2}\right) } \nonumber \\
 & & \qquad \times \determ{\tvval}^{-\half(3\lsd-\rk+1)} \exp\left( -\half \lgtv\pinv \priorscalematrixrank{\rk} \right)    ,
\end{IEEEeqnarray}
%
in which $\lgtv\pinv$ denotes the Moore-Penrose pseudoinverse of $\lgtv$. A prior estimate of $\lgtv$ is $\frac{1}{\rk}\priorscalematrixrank{\rk}$. Thus, in order to maintain consistency between the different ranks, we should set $\priorscalematrixrank{\rk} = \rk \priorscalematrixbase$, where $\priorscalematrixbase$ is an estimate of $\lgtv$.

@@

\subsubsection{Likelihood and Subspace Constraints}
There is no transition density over $\reals^{\lsd}$ for the degenerate model. However, a density does exist with respect to an appropriate measure on the reachable subspace, which provides us with a likelihood function \cite{Diaz-Garcia2006}. We can derive this from the underlying state transition equation,
%
\begin{IEEEeqnarray}{rCl}
 \ls{\ti} &=& \lgtm \ls{\ti-1} + \tvrot \tvfull\msqrt \tn{\ti} \nonumber
 \end{IEEEeqnarray}
 \begin{IEEEeqnarray}{rCl}
 \Rightarrow \tvrot\tr(\ls{\ti}-\lgtm\ls{\ti-1}) &\sim& \normaldist{\zmat}{\tvfull} \nonumber
\end{IEEEeqnarray}
\begin{IEEEeqnarray}{l}
 \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) \propto \determ{\tvfull}^{-\half}  \\
 \times \exp\left(-\half (\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \tvfull\inv \tvrot\tr (\ls{\ti}-\lgtm\ls{\ti-1}) \right) \nonumber     .
\end{IEEEeqnarray}
%
The subspace over which this is defined is,
%
\begin{IEEEeqnarray}{rCl}
 \{\ls{\ti} : \ls{\ti} = \lgtm \ls{\ti-1} + \tvrot z, z \in \reals^{\rk}\}     .
\end{IEEEeqnarray}

The likelihood function for the entire state trajectory is thus,
%
\begin{IEEEeqnarray}{rCl}
 \IEEEeqnarraymulticol{3}{l}{ \den(\ls{1:\timax}|\lgtm, \lgtv) = \den(\ls{0}) \prod_{t=2}^{\timax} \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) } \nonumber \\
 &\propto& \determ{\tvfull}^{-\half(\timax-1)} \exp\Bigg( -\half \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \nonumber \\
 && \qquad\qquad\qquad\qquad\qquad\qquad\qquad \tvfull\inv \tvrot\tr (\ls{\ti}-\lgtm\ls{\ti-1}) \Bigg) \nonumber \\
 %&=& \determ{\tvfull}^{-\half(\timax-1)} \exp\left(-\half\trace\left[ \tvfull\inv \tvrot\tr \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})(\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \right]\right) \nonumber \\
 %&=& \determ{\tvfull}^{-\half\suffstats{0}} \exp\left(-\half\trace\left[ \tvfull\inv \left( \tvrot\tr \lgtm \suffstats{1} \lgtm\tr \tvrot - \tvrot\tr \lgtm \suffstats{2}\tr \tvrot - \tvrot\tr \suffstats{2} \lgtm\tr \tvrot + \tvrot\tr \suffstats{3} \tvrot \right) \right]\right) \\
 &=& \determ{\tvfull}^{-\half\suffstats{0}} \exp\bigg(-\half\trace \big[ \tvfull\inv ( \lgtmrot \suffstats{1} \lgtmrot\tr - \lgtmrot \suffstats{2}\tr \tvrot \nonumber \\
 & & \qquad\qquad\qquad\qquad\qquad -\: \tvrot\tr \suffstats{2} \lgtmrot\tr + \tvrot\tr \suffstats{3} \tvrot ) \big] \bigg)
\end{IEEEeqnarray}
%
in which the sufficient statistics are as before and,
%
\begin{equation}
 \lgtmrot = \tvrot\tr \lgtm      .
\end{equation}

@@

\subsubsection{Sampling the Conditional}
Using the constrained likelihood function and the transformed prior, we now obtain the required posterior conditional distribution $\postden(\lgtmrot,\tvfull|\lgtmorth,\tvrot,\ls{1:\timax})$. This follows the same algebraic steps as in the full rank case and leads to a similar matrix normal--inverse Wishart distribution,
%
\begin{align}
 \tvfull | \tvrot, \ls{1:\timax} &\sim \iwishartdist{\postdof}{\postscalematrix} \\
 \lgtmrot | \tvfull, \tvrot, \ls{1:\timax} &\sim \matrixnormaldist{\postmeanmatrix}{\lgtv}{\postcolumnvariance}     ,
\end{align}
%
with the following updated hyperparameters,
%
\begin{IEEEeqnarray}{rCl}
 \postcolumnvariance\inv                 &=& \priorcolumnvariance\inv + \suffstats{1} \\
 \postmeanmatrix \postcolumnvariance\inv &=& \tvrot\tr\left( \priormeanmatrix \priorcolumnvariance\inv + \suffstats{2} \right) \\
 \postdof                                &=& \rk + \suffstats{0} \\
 \postscalematrix                        &=& (\tvrot\tr\priorscalematrix\inv\tvrot)\inv + \tvrot\tr\left( \suffstats{3} + \priormeanmatrix \priorcolumnvariance\inv \priormeanmatrix\tr \right)\tvrot \nonumber \\
 & & \qquad\qquad\qquad\qquad\qquad -\: \postmeanmatrix \postcolumnvariance\inv \postmeanmatrix\tr
\end{IEEEeqnarray}

@@

\section{Reversible Jump for Rank Learning} \label{sec:rank_learning}

@@

The acceptance probability for this move and its reverse respectively are \cite{Green1995},
%
\begin{IEEEeqnarray}{rCl}
 \mhap(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew) &=& \min\left\{1, \beta \right\} \\
 \mhap(\lgtv\mcnew\to\lgtv,\eval{\rk+1},\evec{\rk+1}) &=& \min\left\{1, \beta\inv \right\} 
\end{IEEEeqnarray}
%
\begin{IEEEeqnarray}{rCl}
 \beta &=& \frac{ \postden(\lgtm, \lgtv\mcnew, \rk+1) \jac(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew) }{ \postden(\lgtm,\lgtv,\rk) \ppslden{\eval{}}(\eval{\rk+1}|\lgtv) \ppslden{\evec{}}(\evec{\rk+1}|\lgtv) } \nonumber \\
 & & \qquad\qquad\qquad\qquad\qquad \times \frac{P(\rk+1\to\rk)}{P(\rk\to\rk+1)}     ,
\end{IEEEeqnarray}
%
where $\jac(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew)$ is the Jacobian of the transformation and $P(\rk+1\to\rk)$ is the probability of performing a move which changes the rank from $\rk$ to $\rk+1$.

The Jacobian of the eigendecomposition for a positive semi-definite matrix is \cite{Diaz-Garcia2006},
%
\begin{IEEEeqnarray}{rCl}
 \jac(\tvval,\tvvec\to\lgtv) &=& 2^{-\rk} \times \prod_{i=1}^{\rk}\eval{i}^{\lsd-\rk} \times \prod_{j=1}^{r}\prod_{i=1}^{j-1}(\eval{i}-\eval{j})      .
\end{IEEEeqnarray}
%
Repeating this for $\lgtv\mcnew$ and taking the ratio of the two, we obtain,
%
\begin{IEEEeqnarray}{l}
 \jac(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew) \nonumber \\
 \qquad\qquad = 2^{-1} \eval{\rk+1}^{\lsd-\rk-1} \prod_{i=1}^{\rk}\eval{i}^{-1} \prod_{i=1}^{\rk}(\eval{i}-\eval{\rk+1})     .
\end{IEEEeqnarray}

\section{Simulations} \label{sec:simulations}

@@

The hyperparameters were set to generally uninformative values. For the full rank model,
%
\begin{IEEEeqnarray}{rClCrCl}
 \priordof &=& \lsd - 1 & \qquad & \priorscalematrix &=& \priordof \idmat \nonumber \\
 \priormeanmatrix &=& \zmat & \qquad & \priorcolumnvariance &=& 100 \idmat      .
\end{IEEEeqnarray}
%
Similarly, for the degenerate model,
%
\begin{IEEEeqnarray}{rClCrCl}
 \priorscalematrixbase &=& \idmat & \qquad & \priormeanmatrix &=& \zmat \nonumber \\
 \priorcolumnvariance &=& 100 \idmat & \qquad & \priortypval &=& 1 \nonumber \\
 \priorrankweight{i} &=& \frac{1}{\lsd}     .
\end{IEEEeqnarray}

@@

\begin{figure}
 \centering
 \subfloat[]{\includegraphics[width=0.45\columnwidth]{figures/toy-basic-R-acf.pdf} \label{subf:toy-R-acf:basic}}
 \subfloat[]{\includegraphics[width=0.45\columnwidth]{figures/toy-degenerate-R-acf.pdf} \label{subf:toy-R-acf:degenerate}}
 \caption{MCMC autocorrelation plots for $\xi_{\ob{}}$ for the full rank transition model \protect\subref{subf:toy-R-acf:basic} and the degenerate transition model \protect\subref{subf:toy-R-acf:degenerate}.}
 \label{fig:toy-R-acf}
\end{figure}

@@

The hyperparameters are set to the same uninformative values as for the toy model, with the following exception,
%
\begin{IEEEeqnarray}{rCl}
 \priorscalematrix &=& 0.001 \priordof \idmat \nonumber \\
 \priorscalematrixbase &=& 0.001 \idmat \nonumber \\
 \priortypval &=& 0.01     .
\end{IEEEeqnarray}

@@

\bibliographystyle{IEEEtran}
\bibliography{/home/pete/Dropbox/PhD/bibliographies/Cleanbib}
% \bibliography{/home/pete/Dropbox/PhD/bibliographies/OTbib}
\end{document}
